---
title: "Bradypus GLM - My First SDM"
author: "Abdul Munawar Mohammed"
date: 06-05-2026
format: html
editor: visual
---

## Introduction

This notebook builds a Species Distribution Model (SDM) for *Bradypus variegatus* (three-toed sloth) using a Generalised Linear Model (GLM). Following the standard protocol by Zurell et al. (2020), we use climate data from WorldClim to predict habitat suitability across South America.

## Objectives

1.  Load and explore species occurrence data
2.  Generate pseudo-absence points
3.  Extract environmental predictors
4.  Check multicollinearity using VIF
5.  Fit and evaluate a logistic GLM
6.  Map predicted habitat suitability

## Species

**Bradypus variegatus** (three-toed sloth) is a tropical mammal native to Central and South America. It thrives in warm, humid environments and is strongly associated with tropical forest biomes.

```{r}
#| label: load-libraries
#| message: false
#| warning: false

library(dismo)
library(terra)
library(ggplot2)
library(dplyr)

```

```{r}
#| label: load-data

# Load Bradypus occurrence data
file <- paste(system.file(package="dismo"), "/ex/bradypus.csv", sep="")
bradypus <- read.csv(file)

# Quick look at the data
head(bradypus)
nrow(bradypus)
```

```{r}
#| label: load-environment
#| message: false

# Load the environmental raster layers
files <- list.files(path=paste(system.file(package="dismo"), "/ex", sep=""),
                    pattern="grd", full.names=TRUE)

predictors <- rast(files)

# See what variables we have
names(predictors)


```

```{r}

#| label: plot-occurrences

# Plot first layer - annual mean temperature
plot(predictors[[1]], main="Bradypus variegatus occurrences")

# Add sloth points in red on top
points(bradypus[,2:3], col="red", pch=20, cex=0.8)

```

```{r}


#| label: pseudo-absences
#| message: false

# Convert to old raster format so dismo can read it
predictors_raster <- as(predictors, "Raster")

# Set a seed so results are reproducible
set.seed(42)

# Generate 116 random absence points
absence <- randomPoints(predictors_raster, n=116)

# Convert to dataframe
absence <- as.data.frame(absence)
colnames(absence) <- c("lon", "lat")

# Quick look
head(absence)
nrow(absence)


```

```{r}

#| label: extract-values
#| message: false

# Extract environmental values at presence points
presence_vals <- extract(predictors, bradypus[,2:3])

# Extract environmental values at absence points  
absence_vals <- extract(predictors, absence)

# Add a column to label them 1 (present) or 0 (absent)
presence_vals$presence <- 1
absence_vals$presence <- 0

# Combine into one dataset
model_data <- rbind(presence_vals, absence_vals)

# Quick look
head(model_data)
nrow(model_data)

```

```{r}

#| label: clean-data

# Remove any rows with missing values
model_data <- na.omit(model_data)

# Check final size
nrow(model_data)

# Check balance of presence vs absence
table(model_data$presence)

```

```{r}

#| label: split-data

set.seed(42)

# Create a random 80/20 split
train_index <- sample(1:nrow(model_data), size = 0.8 * nrow(model_data))

train_data <- model_data[train_index, ]
test_data  <- model_data[-train_index, ]

nrow(train_data)
nrow(test_data)

```

```{r}


#| label: vif-check
#| message: false
#| warning: false

library(car)

# Remove highly correlated variables and retest
glm_vif2 <- glm(presence ~ bio17 + bio16 + biome,
                data = train_data,
                family = binomial(link = "logit"))

vif_values2 <- vif(glm_vif2)

vif_table2 <- data.frame(
  Variable = names(vif_values2),
  VIF = round(vif_values2, 2)
)

vif_table2 <- vif_table2[order(vif_table2$VIF, decreasing = TRUE), ]
print(vif_table2)

```

```{r}
#| label: fit-glm

# Refit GLM with VIF-selected variables only
glm_model <- glm(presence ~ bio16 + bio17 + biome,
                 data = train_data,
                 family = binomial(link = "logit"))

summary(glm_model)

```

```{r}

#| label: evaluate-model
#| message: false
#| warning: false

library(pROC)

# Predict on test data
predictions <- predict(glm_model, newdata = test_data, type = "response")

# Calculate AUC
roc_curve <- roc(test_data$presence, predictions)
auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, main="ROC Curve - Bradypus GLM", col="blue", lwd=2)

```

```{r}

#| label: effect-size-plot
#| message: false

# Extract coefficients from the model
coefs <- summary(glm_model)$coefficients
effect_sizes <- data.frame(
  Variable = rownames(coefs)[-1],
  Effect_Size = coefs[-1, "Estimate"],
  Std_Error = coefs[-1, "Std. Error"]
)

# Plot effect sizes with error bars
ggplot(effect_sizes, aes(x = reorder(Variable, Effect_Size), y = Effect_Size)) +
  geom_bar(stat = "identity", fill = "#11aa96") +
  geom_errorbar(aes(ymin = Effect_Size - Std_Error, 
                    ymax = Effect_Size + Std_Error), width = 0.2) +
  coord_flip() +
  labs(title = "Effect Size of Each Variable",
       x = "Variable",
       y = "Coefficient Estimate") +
  theme_minimal()


```

```{r}

#| label: response-curves
#| message: false

# Response curve for bio16
bio16_range <- seq(min(train_data$bio16), max(train_data$bio16), length.out = 100)
response_data <- data.frame(
  bio16 = bio16_range,
  bio17 = mean(train_data$bio17),
  biome = mean(train_data$biome)
)
response_data$predicted <- predict(glm_model, newdata = response_data, type = "response")

ggplot(response_data, aes(x = bio16, y = predicted)) +
  geom_line(color = "#378ADD", linewidth = 1) +
  labs(title = "Species Response Curve — bio16 (Wettest Quarter Rainfall)",
       x = "Precipitation of Wettest Quarter (mm)",
       y = "Predicted Probability of Presence") +
  theme_minimal()

```

```{r}

```

```         
```

## Predicted Bradypus Habitat Suitability Map  For SA

```{r}

#| label: interactive-map
#| message: false
#| warning: false

library(leaflet)
library(terra)

# Convert prediction map to a format leaflet can read
prediction_map <- predict(predictors, glm_model, type="response")

# Define colour palette
pal <- colorNumeric(
  palette = c("white", "yellow", "orange", "red"),
  domain = values(prediction_map),
  na.color = "transparent"
)

# Create interactive map
leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addRasterImage(prediction_map, 
                 colors = pal, 
                 opacity = 0.7,
                 project = TRUE) %>%
  addLegend(pal = pal,
            values = values(prediction_map),
            title = "Habitat suitability",
            position = "bottomright") %>%
  addScaleBar(position = "bottomleft")

```
